---
title: 'Module 10: rstan'
output:
  pdf_document: default
  html_notebook: default
---


To learn about rstan, consider this vignette: 

https://cran.r-project.org/web/packages/rstan/vignettes/rstan.html

See here for prior choice recommendation:

https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations

In this notebook, we start simpler, estimating just a mean and variance, followed by the fitting of a hierarchical model. 
We compare the model used to that used (implicitly) in brm.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(brms)

library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

library(tidybayes) # added mostly as a reference for future use, for easier manipulation of posterior samples. 
# used here for function `point_interval'
```


# Radon data
Read in the radon data and process (copied from earlier module)
```{r}
# house level data
d <- read.table(url("http://www.stat.columbia.edu/~gelman/arm/examples/radon/srrs2.dat"), 
                header=T, sep=",")

# deal with zeros, select what we want, make a fips (county) variable to match on 
d <- d %>% 
  mutate(activity = ifelse(activity==0, 0.1, activity)) %>% 
  mutate(fips = stfips * 1000 + cntyfips) %>%   
  dplyr::select(fips, state, county, floor, activity)

# county level data
cty <- read.table(url("http://www.stat.columbia.edu/~gelman/arm/examples/radon/cty.dat"), 
                  header = T, sep = ",")
cty <- 
  cty %>% 
  mutate(fips = 1000 * stfips + ctfips) %>% 
  dplyr::select(fips, Uppm) %>%
  rename(ura_county = (Uppm))

dmn <- d %>% 
  filter(state=="MN") %>% # Minnesota data only
  dplyr::select(fips, county, floor, activity) %>% 
  left_join(cty) 

dat <- 
  dmn%>%
  mutate(y = log(activity), log_ur = log(ura_county))
head(dat)
```



# Using rstan to estimate mean and variance 

Start simple, let $y_i =$ log(radon), and assume 
$$y_i|\mu, \sigma \sim N(\mu, \sigma^2),$$
to estimate $\mu$ and $\sigma$.

We need to define data inputs and a stan model file.

For the stan model file template, go to "File", "new file", "stan file".
Note that the default is to estimate $\mu$ and $\sigma$ as in our seting!
The default template has flat priors, we can add priors.
Save it, I called it "module10_mean_addpriors.stan".

Create data inputs: 
```{r}
stan_dat <- list(y = dat$y, N = length(dat$y))
```


```{r, results = "hide"}
fit <- stan(file = 'module10_stan_mean_addpriors.stan', data = stan_dat)
```

Showing some default oupputs, see also the Rstan vignette

```{r}
print(fit)
```


```{r}
plot(fit)
```

```{r}
pairs(fit, pars = c("mu", "sigma"))
```

```{r}
pairs(fit, pars = c("mu", "sigma"))
```


```{r}
traceplot(fit, pars = c("sigma"))
```



## Compare to brm-based model

```{r, results = "hide"}
fit_brm <- brm(y ~ 1, data = dat,
        chains = 4, iter = 1000, warmup = 500, cores = getOption("mc.cores", 4))
```


```{r}
summary(fit_brm)
```

```{r}
print(fit)
```

```{r}
stan_data_from_brm <- standata(fit_brm)
names(stan_data_from_brm)
# N and y (called Y here) are same as in stan_dat
stan_dat$N
stan_data_from_brm$N

```

```{r}
summary(stan_data_from_brm$Y)
summary(dat$y)
```

```{r}
stancode(fit_brm)
```


See slides re what's going on in this model!

For just the priors
```{r}
get_prior(y~1, data = dat)
```

# Fit a multilevel regression model 

Consider the following model: $y_i|\mu_i, \sigma \sim N(\mu_i, \sigma^2),$ where
$$\mu_i = \mu_{\alpha} + \eta_{j[i]} + \beta x_i,$$
where $j[i]$ refers to county and $x_i$ to floor indicator.

The stan model for this is given in module10_hier_regression.stan. 

```{r}
dat <- dat %>%
  mutate(county_id = as.numeric(as_factor(county)))
stan_dat2 <- list(y = dat$y, N = length(dat$y), x = dat$floor, county_id = dat$county_id, 
                  J = max(dat$county_id))
```


```{r, results = "hide"}
fit2 <- stan(file = 'module10_hier_regression.stan', data = stan_dat2)
```

```{r}
print(fit)
```


